---
title: "Read Me"
output: md_document
---

semtree
======

```{r echo=FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE
)
```

[![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.1116294.svg)](https://doi.org/10.5281/zenodo.1116294)
[![cran version](http://www.r-pkg.org/badges/version/semtree)](https://cran.r-project.org/package=semtree)
[![rstudio mirror downloads](http://cranlogs.r-pkg.org/badges/semtree)](https://github.com/r-hub/cranlogs.app)
<!-- badges: start -->
[![R-CMD-check](https://github.com/brandmaier/semtree/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/brandmaier/semtree/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

![Code size](https://img.shields.io/github/languages/code-size/brandmaier/semtree.svg)
![contributions](https://img.shields.io/badge/contributions-welcome-brightgreen.svg?style=flat)
[![License: GPL v3](https://img.shields.io/badge/License-GPLv3-blue.svg)](https://www.gnu.org/licenses/gpl-3.0)


## What is this?

An R package for estimating Structural Equation Model (SEM) Trees and Forests. They are a fusion of SEM and decision trees, or SEM and random forests respectively. While SEM is a confirmatory modeling technique, SEM trees and forests allow to explore whether there are predictors that provide further information about an initial, theory-based model. Potential use cases are the search for potential predictors that explain individual differences, finding omitted variables in a model, or exploring measurement invariance over a large set of predictors. A recent overview is in our latest book chapter in the SEM handbook (Brandmaier & Jacobucci, 2023). 

## Install

Install the latest stable version from CRAN:
```{r eval=FALSE}
install.packages("semtree")
```

To install the latest semtree package directly from GitHub, copy the following line into R:
```{r, eval=FALSE}
library(devtools)
devtools::install_github("semtree/brandmaier")

# even better: install with package vignette (extra documentation)
devtools::install_github("brandmaier/semtree",force=TRUE, build_opts = c())
```

## Usage

Package documentation and use-cases with runnable R code can be found on our github pages: [https://brandmaier.github.io/semtree/](https://brandmaier.github.io/semtree/).

You may also want to visit the semtree website: [https://brandmaier.de/semtree](https://brandmaier.de/semtree)

Package vignettes (shipped with the package) contain documentation on how to use the package. Simply type this in R once you have loaded the package:

```{r eval=FALSE}
browseVignettes("semtree")
```

## References

Theory and method:

- Brandmaier, A. M., & Jacobucci, R. C. (2023). Machine-learning approaches to structural equation modeling. In R. H. Hoyle (Ed.), Handbook of structural equation modeling (2nd rev. ed., pp. 722–739). Guilford Press.

- Arnold, M., Voelkle, M.C., and Brandmaier, A.M. (2021). Score-guided structural equation model trees. _Frontiers in psychology_, 11, 564403.

- Brandmaier, A. M., Driver, C., & Voelkle, M. C. (2019). Recursive partitioning in continuous time analysis. In K. van Montfort, J. Oud, & M. C. Voelkle (Eds.), Continuous time modeling in the behavioral and related sciences. New York: Springer.

- Brandmaier, A. M., Prindle, J. J., McArdle, J. J., & Lindenberger, U. (2016). Theory-guided exploration with structural equation model forests. _Psychological Methods_, 21, 566-582. \doi{doi:10.1037/met0000090}

- Brandmaier, A. M., von Oertzen, T., McArdle, J. J., & Lindenberger, U. (2014). Exploratory data mining with structural equation model trees. In J. J. McArdle & G. Ritschard (Eds.), Contemporary issues in exploratory data mining in the behavioral sciences (pp. 96-127). New York: Routledge.

- Brandmaier, A. M., von Oertzen, T., McArdle, J. J., & Lindenberger, U. (2013). Structural equation model trees. _Psychological Methods_, 18, 71-86. \doi{doi:10.1037/a0030001}

Applied examples (there are many more):

Brandmaier, A. M., Ram, N., Wagner, G. G., & Gerstorf, D. (2017). Terminal decline in well-being: The role of multi-indicator constellations of physical health and psychosocial correlates. Developmental Psychology.
