---
title: "SEM Forests"
author: "Andreas Brandmaier"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SEM Forests}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(semtree)
```

This example demonstrates how SEM forests can be grown. SEM forests are ensembles of typically hundreds to thousands of SEM trees. Using permutation-based variable importance estimates, we can aggregate the importance of each predictor for improving model fit. 

Here, we use the `affect` dataset and a simple SEM with only a single observed variable and no latent variables.

## Load data

Load affect dataset from the `psychTools` package. These are data from two studies conducted in the Personality, Motivation and Cognition Laboratory at Northwestern University to  study affect dimensionality and the relationship to various personality dimensions.

```{r}
library(psychTools)
data(affect)

affect <- affect[complete.cases(affect$state2), ]
affect$BDI <- NULL # Drop affect because it is NA
affect$Study <- NULL # Drop Study because it is constant

knitr::kable(head(affect))

affect$Film <- as.factor(affect$Film)
affect$lie <- as.ordered(affect$lie)
affect$imp <- as.ordered(affect$imp)
```

## Create simple model of state anxiety

The following code implements a simple SEM with only a single manifest variables and two parameters, the mean of state anxiety after having watched a movie (`state2`), $\mu$, and the variance of state anxiety, $\sigma^2$.

```{r}
library(OpenMx)
manifests<-c("state2")
latents<-c()
model <- mxModel("Univariate Normal Model", 
type="RAM",
manifestVars = manifests,
latentVars = latents,
mxPath(from="one",to=manifests, free=c(TRUE), 
       value=c(50.0) , arrows=1, label=c("mu") ),
mxPath(from=manifests,to=manifests, free=c(TRUE), 
       value=c(100.0) , arrows=2, label=c("sigma2") ),
mxData(affect, type = "raw")
);

result <- mxRun(model)
```

These are the estimates of the model when run on the entire sample:

```{r}
summary(result)
```

## Forest

Create a forest control object that stores all tuning parameters of the forest. Note that we use only 5 trees for illustration. Please increase the number in real applications to several hundreds. To speed up computation time, consider score-based test for variable selection in the trees.

```{r}
control <- semforest_score_control(num.trees = 50)
print(control)
```

Now, run the forest using the `control` object:

```{r precompute, echo=FALSE, eval=FALSE}
cat(getwd())
cache_file <- "inst/extdata/cache.rds"
if (!file.exists(cache_file)) {
forest <- semforest( model=model,
                     data = affect, 
                     control = control,
                     covariates = c("Film", "state1",
                                    "PA2","NA2","TA2"))
saveRDS(object=forest, file=cache_file)
} else {
  forest <- readRDS(cache_file)
}

```

```{r forest, eval=TRUE, message=FALSE, echo=TRUE, warning=FALSE, results="hide"}
forest <- semforest( model=model,
                     data = affect, 
                     control = control,
                     covariates = c("Film", "state1",
                                    "PA2","NA2","TA2"))
```

## Variable importance

Next, we compute permutation-based variable importance. This may take some time.

```{r varimp}
vim <- varimp(forest)
print(vim, sort.values=TRUE)
plot(vim)
```

From this, we can learn that variables such as `NA2` representing negative affect (after the movie), `TA2` representing tense arousal (after the movie), and `state1` representing the state anxiety before having watched the movie, are the best predictors of difference in the distribution of state anxiety (in either mean, variance or both) after having watched the movie. 

Variable importance estimates are averaged across all trees by taking the mean importance of each variable over all trees. Sometimes, these estimates may be distorted if a single tree produces unstable estimates. To obtain a more robust estimate, it may make sense to take the median across all trees to obtain the variable importance estimates:

```{r}
print(vim, sort.values=TRUE, aggregate="median")
plot(vim, aggregate="median")
```

To inspect, how well these estimates converged, we can look at the convergence plot that shows the estimates as a function of the number of trees considered:

```{r}
plot(vim, convergence=TRUE)
```

# Partial Dependence

Partial dependence plots visualize the relationship of a predictor and a selected model parameter. This is a powerful tool to help with interpreting SEM forests. 

```{r pdp}
pdp <- partialDependence(forest,reference.var = "TA2")
```

```{r plotpdp}
plot(pdp, parameter="mu")
```
