---
title: "Getting Started with the semtree package"
author: "Andreas M. Brandmaier"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with the semtree package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Load the Package

We first load the `semtree` package and the `OpenMx` package for specifying our SEM.

```{r setup}
library(semtree)
library(OpenMx)
```

## Simulate data

Now, we simulate some data from a linear latent growth curve model (that is, a random intercept and random slope over time). The dataset will be called `growth.data`. The dataset contains five observations for each individual (`X1` to `X5`) and one predictor `P1`. The predictor is dichotomous and predicts a (quite large) difference in mean slope.

```{r simdata}
set.seed(23)
N <- 1000
M <- 5
icept <- rnorm(N, 10, sd = 4)
slope <- rnorm(N, 3, sd = 1.2)
p1 <- sample(c(0, 1), size = N, replace = TRUE)
loadings <- 0:4
x <-
  (slope + p1 * 5) %*% t(loadings) + 
  matrix(rep(icept, each = M), byrow = TRUE, ncol = M) + 
  rnorm(N * M, sd = .08)
growth.data <- data.frame(x, factor(p1))
names(growth.data) <- c(paste0("X", 1:M), "P1")
```

## Specify an OpenMx model

Now, we specify a linear latent growth curve model using OpenMx's path specification. The model has five observed variables. Residual variances are assumed to be identical over time.

```{r}
manifests <- names(growth.data)[1:5]
growthCurveModel <- mxModel("Linear Growth Curve Model Path Specification",
    type="RAM",
       manifestVars=manifests,
    latentVars=c("intercept","slope"),
    mxData(growth.data, type="raw"),
    # residual variances
    mxPath(
        from=manifests,
        arrows=2,
        free=TRUE,
        values = c(.1, .1, .1, .1, .1),
        labels=c("residual","residual","residual","residual","residual")
    ),
    # latent variances and covariance
    mxPath(
        from=c("intercept","slope"),
        arrows=2,
        connect="unique.pairs",
        free=TRUE,
        values=c(2, 0, 1),
        labels=c("vari", "cov", "vars")
    ),
    # intercept loadings
    mxPath(
        from="intercept",
        to=manifests,
        arrows=1,
        free=FALSE,
        values=c(1, 1, 1, 1, 1)
    ),
    # slope loadings
    mxPath(
        from="slope",
        to=manifests,
        arrows=1,
        free=FALSE,
        values=c(0, 1, 2, 3, 4)
    ),
    # manifest means
    mxPath(
        from="one",
        to=manifests,
        arrows=1,
        free=FALSE,
        values=c(0, 0, 0, 0, 0)
    ),
    # latent means
    mxPath(
        from="one",
        to=c("intercept", "slope"),
        arrows=1,
        free=TRUE,
        values=c(1, 1),
        labels=c("meani", "means")
    )
) # close model

# fit the model to the entire dataset
growthCurveModel <- mxRun(growthCurveModel)
```

## Run a tree

Now, we grow a SEM tree using the `semtree` function, which takes the model and the dataset as input. If not specified otherwise, SEM tree will assume that all variables in the dataset, which are not observed variables in the dataset are potential predictors.

```{r message=FALSE,warning=FALSE,results="hide"}
tree <- semtree(model = growthCurveModel, 
                data = growth.data)
```

## Plotting

Once the tree is grown, we can plot it:

```{r}
plot(tree)
```