---
title: "Score-based Tests"
author: "Andreas M. Brandmaier"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{score-based-tests}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```



## Load data

Load affect dataset from the `psychTools` package. These are data from two studies conducted in the Personality, Motivation and Cognition Laboratory at Northwestern University to  study affect dimensionality and the relationship to various personality dimensions.

```{r}
library(psychTools)
data(affect)

affect$Film <- as.ordered(affect$Film)


tree.data <- affect[,c("state1","Film","neur","ext")] 

knitr::kable(head(tree.data))
```

## Create simple model of state anxiety

```{r}
library(OpenMx)
manifests<-c("state1")
latents<-c()
model <- mxModel("Simple Model", 
type="RAM",
manifestVars = manifests,
latentVars = latents,
mxPath(from="one",to=manifests, free=c(TRUE), value=c(1.0) , arrows=1, label=c("mu") ),
mxPath(from=manifests,to=manifests, free=c(TRUE), value=c(1.0) , arrows=2, label=c("sigma2") ),
mxData(tree.data, type = "raw")
);

result <- mxRun(model)
summary(result)
```

## Score-based Tests

Use score-based tests to create the tree.

```{r}
library(semtree)
ctrl = semtree.control(method="score")
#tree = semtree(model=result, data=tree.data, control=ctrl)
```

Plot the tree

```{r}
plot(tree)
```